---
title: "Modeling Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling Setup}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: ref.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/",
  fig.align = "center",
  fig.dim = c(8, 6), 
  out.width = "75%"
)
library("Rprobit")
```

This vignette explains how the model can be set up and how the data is
included. We use the Train data set from the {mlogit} package
[@Croissant:2020] for illustration.The data structure is analogous to
the {RprobitB} package.

## Requirements to choice data

**Rprobit** helps in modeling the choice of individual deciders of one
alternative from a finite set of choice alternatives. This choice set
has to fulfill three properties [@Train:2009]: Choices need to be

1.  mutually exclusive (one can choose one and only one alternative that
    are all different),

2.  exhaustive (the alternatives do not leave other options open),

3.  and finitely many.

Every decider may take one or repeated choices (called choice
occasions). The data set thus contains information on

1.  an identifier for each decider (and optionally for each choice
    situation),

2.  the choices,

3.  alternative and decider specific covariates.

Additionally, {Rprobit} asks the following formal requirements:

1.  The data set **must** be in "wide" format, that means each row
    provides the full information for one choice occasion.[^1]

2.  It **must** contain a column with unique identifiers for each
    decision maker. Additionally, it **can** contain a column with
    identifier for each choice situation of each decider. If this
    information is missing, these identifier are generated automatically
    by the appearance of the choices in the data set.[^2]

3.  It **can** contain a column with the observed choices. Such a column
    is required for model fitting but not for prediction.

4.  It **must** contain columns for the values of each alternative
    specific covariate for each alternative and for each decider
    specific covariate.

[^1]: The {tidyr} package [contains
    functionality](https://tidyr.tidyverse.org/articles/pivot.html) that
    can transform a `data.frame` into this format.

[^2]: The choice situation identifier are irrelevant for model
    estimation because {RprobitB} does not model correlation over choice
    occasions. They are useful to identify specific choice occasions
    later on.

### Example

The Train data set contains 2929 stated choices by 235 Dutch individuals
deciding between two virtual train trip options based on the price, the
travel time, the level of comfort, and the number of changes. It
fulfills the above requirements: Each row represents one choice
occasion, the columns `id` and `choiceid` identify the deciders and the
choice occasions, respectively. The column `choice` gives the observed
choices. Four alternative-specific covariates are available, namely
`price`, `time`, `change`, and `comfort`. There values are given for
each alternative.[^3]

[^3]: For alternative specific variables, the alternative names must be
    added to the covariates via the `_` separator.

```{r, overview-Train-data}
data("Train", package = "mlogit")
str(Train)
```

## The setup function

The central object for {Rprobit} is the `Rprobit_obj` object. It stores
all information on the data, the specified model and the estimation. In
the course of using {Rprobit} the object is changed and enhanced with
features.

The object contains the following fields (not all need to be specified
at initialization): 

- data: \code{\link{data}} object 
- data_raw: \code{\link{data_raw_cl}} object: a wide representation of choice data, see
above. 
- form: \code{\link{form}} object: model formula 
- re: vector of strings listing the names of random effects regressors 
- mod: \code{\link{mod_cl}} object: see model specification 
- vars: vector of strings of variable names 
- alt_names: vector of strings of names of choice alternatives 
- theta: \code{\link{theta}} object; estimated parameter vector 
- theta_0: \code{\link{theta}} object; true parameter vector used for data generation in simulated context. 
- ll real: log likelihood value 
- H real matrix: Hessian at estimate 
- J real matrix: variance of score at estimate 
- grad: real vector; gradient at theta 
- fit: real; information on the fit
- vv: real matrix; covariance of estimated parameters 
- control: list of parameters controlling estimation 
- info: providing information of estimation.

Many of these fields are supplied by specific functions. There are two
main ways to generate an `Rprobit_obj` object:

1.  Providing a raw data set and a model.
2.  Providing a system within a model and the specification of a data
    set in a simulation setting.

### Setting up the object from data

The `Rprobit_obj` can be initialized using a data set in wide format. In
addition in this case the model must be supplied in the form of either
(A) a `mod` object or (B) in form of a formula which can be accompanied
by

1.  `re` (optional: default is empty): a list providing the names of
    variables whose coefficients are random
2.  `corr_re` (optional: default is FALSE): boolean whether correlations
    of the mixed coefficients should be included in the model
3.  `error_struc` (optional: default is `"full"`): string indicating the
    error structure. Choices here are `"diag"`, `"fixed"`(to 0.5 times
    the identity matrix), `"ident"` (the identity matrix) and `"full"`

In that case `setup_Rprobit` converts the raw data from the wide format
to the unified format (see model definition) and extracts the number of
deciders, the number of choice occasions per decider from the data. In
case of (B) also a `mod` object is generated.

**Example**
For the Train data set from the package {mlogit} an {Rprobit} object can be set up using a formula implying that only
type 1 variables are included. 

```{r, setup}
require(mlogit)
data("Train", package = "mlogit")
Train$choiceid <- 1:nrow(Train)
train_mod <- setup_Rprobit(form = choice ~ price + time + change + comfort | 0 ,  data_raw = Train,  ids = c("id"))
```
The resulting output contains the raw data, the transformed data, the formula and the `mod` object listing 235 deciders facing between 5 and 19 decisions between two alternatives. The model contains four parameters in the vector $\beta$ while $\Omega$ does not exist (equals zero, that is) and $\Sigma$ is fixed to a default value with no parameters:

$$
\Sigma = \left( \begin{array}{cc} 0 & 0 \\ 0 & 1 \end{array} \right)
$$

which in this case is the `full` specification including fixing the scale. 

All remaining fields of the `Rprobit` object are left unspecified. 

### Simulated data set 
The second option for `setup_Rprobit` provides access to simulated data. In this case `form` (a model formula containing the specification of all variables including their type), `mod` (defining the model parametrization; see [Objects](objects.html)) and `control` needs to be specified. This object contains all relevant information on the number of deciders and the number of choice occasions faced by them, the number of alternatives and the names of the variables to be generated. All variables are drawn iid standard normally. 

A parameter vector can be specified, else it is drawn from a standard normal distribution. A seed can be provided such that data generation is reproducible. 

**Example**


```{r, setup-simul}
form <- choice ~ V1 + V2 + V3 | 0 
re <- c("V1","V2")

### Define mod-object 
mod <- mod_cl$new(
  Hb   = diag(3)[,-3],
  fb   = as.matrix(c(0,0,1),col=1),
  HO   = diag(3),
  fO   = matrix(0,3,1),
  HL   = diag(6)[,-c(1,2,3)],
  fL   = matrix(0,6,1),
  ordered = FALSE
)
control_simul <- list(Tp = sample(1:3,1000,replace=TRUE))
Rprobit_obj <- setup_Rprobit(form = form, mod = mod, re = re, seed = 1, control = control_simul)
```

## References
